---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences*
  |  **Examples of Chapter 18:**
  |  **Grouped Multivariate Data**
author: |
  | Kimmo Vehkalahti, Brian S. Everitt; edited by C.-F. Sheu
  |
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  rmarkdown::github_document:
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  fig.align = 'center',
  comment = "",
  tidy = FALSE,     # display code as typed
  size = "small")   # slightly smaller font for code)
```


```{r message=FALSE, warning=FALSE}
# check to see if the pacman package is there
# if not install it and then use it manage packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, survey, ICSNP, Hotelling, multicon)
```

\pagebreak

## Example: Wechsler Adult Intelligence Scale-Revised Test (WAISRT)

Data were collected on 20 male and 20 female right-handed
Anglo psychology students at a large university in the United States. The
subjects took three subtests of the Wechsler Adult Intelligence Scale-Revised
test. The scores recorded were full-scale IQ (FSIQ), 
verbal IQ (VIQ), and performance IQ (PIQ).

## Table 18.1: WAISRT Scores for Men and Women

```{r}
fLoc <- "https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/iq.csv"
iq_scores <- read.csv(fLoc)
```

```{r}
glimpse(iq_scores)
head(iq_scores)
tail(iq_scores)
```


```{r}
# number of men, number of women, number of variables
n1 = 20; n2 = 20; q = 3
# group labels: 20 each
sex <- as.factor(rep(c('M', 'F'), c(n1, n2)))
```

```{r}
# Hotelling's T2 test
# group mean vectors and covariances
m1 <- apply(iq_scores[sex=='M', ], 2, mean)
m2 <- apply(iq_scores[sex=='F', ], 2, mean)
S1 <- var(iq_scores[sex=='M', ])
S2 <- var(iq_scores[sex=='F', ])
# pooled covariance
S <- ((n1-1)*S1 + (n2-1)*S2) / (n1+n2-2)
# Hotelling's T-square
T2 <- t(m1-m2) %*% solve(S) %*% (m1-m2)
# T-square to F
F <- (n1+n2-q-1)*T2/((n1+n2-2)*q)
# F to p-value, one-sided
pvalue <- 1 - pf(F, q, n1+n2-q-1)
# output as list
list(T2=as.numeric(T2), F=as.numeric(F), pvalue=as.numeric(pvalue)) 
```

Quoting our own text on p.367:

Here, Hotelling's $T^2$ takes the value 
`r round(T2, digits = 2)`, 
with the corresponding $F$-statistic being `r round(F, 2)`, 
having 3 and 36 degrees of freedom; the associated $p$-value is 
`r round(pvalue, 2)`. 
There is no evidence of a gender difference on the three measures of IQ.

The following sections do not appear in the textbook.

## HotellingsT2{ICSNP}

```{r}
Y <- as.matrix(iq_scores[,1:3])
ICSNP::HotellingsT2(Y ~ sex)
```

## Data Visualization

```{r fig18.pair, fig.width=5, fig.height=5}
lattice::splom(~iq_scores[,1:3])
```

```{r}
# create gender group ID and subject ID
# convert from wide to long format
iq_scores$Gender <- as.factor(sex)
iq_scores$ID <- as.factor(paste0("S", 101:140))
iqL <- iq_scores %>% gather(key = iq, value = score, 1:3)
```

```{r}
glimpse(iqL)
```

```{r}
ggplot(iqL, aes(Gender, score)) +
 geom_boxplot() +
 facet_grid(. ~ iq) +
 labs(x = "Gender", y = "IQ score") +
 theme_bw()
```

## Multiple t-tests

```{r}
lapply(iq_scores[, c("FSIQ", "VIQ", "PIQ")], 
       function(x) t.test(x ~ iq_scores$Gender, var.equal = TRUE))
```

```{r fig18.dfp}
lapply(iq_scores[, c("FSIQ", "VIQ", "PIQ")], 
       function(x) diffPlot(x ~ iq_scores$Gender, 
                            ylab = "", xlab = "",
                            main = "Plot of Means with Floating Axis for Mean Difference"))
```

\pagebreak

## Reference

Willerman, L., Schultz, R., Rutledge, J.N. \& Bigler, E. D. (1991). 
In vivo brain size and intelligence. *Intelligence*, 15, 223–228. https://doi.org/10.1016/0160-2896(91)90031-8

\pagebreak

## Session information

```{r session_info, include=FALSE}
# set include=TRUE to view the results
sessionInfo()
```