---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  **Examples of Chapter 9:**
  |  **Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables**
author: |
  | "Kimmo Vehkalahti, Brian S. Everitt; edited by C.-F. Sheu"
  |
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  rmarkdown::github_document:
  pdf_document:
    latex_engine: xelatex
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  fig.align = 'center',
  comment = "",
  tidy = FALSE,     # display code as typed
  size = "small")   # slightly smaller font for code)
```


\pagebreak

## Example: Rat weight over time for different diets

Data from a nutrition study conducted in three groups of rats
The three groups were put on different diets, and
each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period. 
The question of most interest is whether the growth profiles of the
three groups differ.

```{r message=FALSE, warning=FALSE}
# check to see if the pacman package is there
# if not install it and then use it manage packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, nlme, lme4, gridExtra)
```

## Table 9.1: Body Weights of Rats 

```{r}
# file location
fLoc <- "https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt"
```

```{r}
# data input
RATS <- read.table(fLoc, header = TRUE, sep = '\t')
```

```{r}
RATS <- within(RATS, {
  ID <- factor(ID)
  Group <- factor(Group)})
```

```{r}
glimpse(RATS)
```

The data set in also available from the *nlme* package
as a grouped data frame object called *BodyWeight*.

```{r}
knitr::kable(head(nlme::BodyWeight, 15))
```


```{r}
RATSL <- nlme::BodyWeight
```

```{r}
names(RATSL) <- c("Weight", "Time", "ID", "Group")
```

```{r}
glimpse(RATSL)
```


\pagebreak

## Figure 9.1

```{r}
# set black and white theme
ot <- theme_set(theme_bw())
```


```{r fig9.1}
ggplot(data = RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_text(aes(label = Group)) + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
```

\pagebreak

## Table 9.3

```{r}
RATS_reg <- lm(Weight ~ Time + Group, data = RATSL)
summary(RATS_reg)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```

\pagebreak

## Figure 9.2

```{r fig9.2}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") + 
  theme(legend.position = "top") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
```

\pagebreak

## Figure 9.3

```{r fig9.3, fig.width = 10, fig.height = 10, fig.align='center'}
pairs(RATS[, 3:13], cex = 0.7)
```

\pagebreak

## Table 9.4

```{r}
RATS_ref <- lme4::lmer(Weight ~ Time + Group + (1 | ID), 
                       data = RATSL, REML = FALSE)
summary(RATS_ref)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```

\pagebreak

## Table 9.5

```{r}
RATS_ref1 <- lme4::lmer(Weight ~ Time + Group + (Time | ID), 
                        data = RATSL, REML = FALSE)
summary(RATS_ref1)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```

```{r}
anova(RATS_ref1, RATS_ref)
```

\pagebreak

## Table 9.6

```{r}
RATS_ref2 <- lme4::lmer(Weight ~ Time * Group + (Time | ID), 
                        data = RATSL, REML = FALSE)
summary(RATS_ref2)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```

```{r}
anova(RATS_ref1, RATS_ref2)
```

\pagebreak

## Figure 9.4

```{r}
# add fitted values to he original data frame 
RATSL <- data.frame(RATSL, Fitted = fitted(RATS_ref2))
```

```{r fig9.4}
# observed
g1 <- ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Weight (grams)") + 
  # "none" in the book
  theme(legend.position = "right") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  ggtitle("Observed")
# fitted
g2 <- ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Weight (grams)") + 
  theme(legend.position = "right") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + 
  ggtitle("Fitted")
# side by side
grid.arrange(g1, g2, nrow = 1)
```

\pagebreak

## Session information

```{r session_info, include=FALSE}
# set include=TRUE to view the results
devtools::session_info()
```