---
title:  |  
  | *Multivariate Analysis for the Behavioral Sciences,*
  | **Examples of Chapter 10:**
  | **Analysis of Longitudinal Data III: Non-Normal Responses**
author: |
  | "Kimmo Vehkalahti, Brian S. Everitt; edited by C.-F. Sheu"
  |
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  rmarkdown::github_document:
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  fig.align = 'center',
  comment = "",
  tidy = FALSE,     # display code as typed
  size = "small")   # slightly smaller font for code)
```

```{r message=FALSE, warning=FALSE}
# check to see if the pacman package is there
# if not install it and then use it manage packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, geepack)
```

\pagebreak

## Example: Beat the Blues Revisited

Longitudinal data from a clinical trial of an interactive, multimedia program known as "Beat the Blues" designed to deliver cognitive behavioral therapy to depressed patients via a computer terminal. Patients with depression recruited in primary care were randomised to either the Beating the Blues program, or to "Treatment as Usual (TAU)". 

A number of outcome measures were used in the trial, but here we concentrate
on the Beck Depression Inventory II (BDI; Beck et al., 1996). Measurements
on this variable were made on the following five occasions:

- Prior to treatment,

- 2, 4, 6, and 8 months after treatment began.

Two additional explanatory variables are also available for each patient: the first, drug, is whether the
patient was taking antidepressant drugs (yes or no), and the second, length,
is the length of the current episode of depression categorized into less than
six months (<6m) or more than six months (>6m). The NAs (not available)
in Table 9.7 indicate where a protocol-specified measurement of the BDI was 
not made; here, all the NAs are due to patients dropping out of the study.
How dropouts might affect the results obtained from the analysis of the data
will be discussed later in the chapter. The main question of interest here is to
estimate the treatment effect of the BtB program.

```{r}
# file location
fLoc <- "https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BtB.txt"
```

```{r}
# data input
BtB <- read.table(fLoc, header = TRUE, sep = '\t')
```

```{r}
# to make sure that the factor levels are logical (esp. Treatment):
BtB <- within(BtB, {
  Drug <- factor(Drug, levels = c("No", "Yes")) # default
  Length <- factor(Length, levels = c("<6m", ">6m")) # default
  Treatment <- factor(Treatment, levels = c("TAU", "BtheB")) # NOT default!
})
```

```{r}
glimpse(BtB); head(BtB); tail(BtB)
```


```{r}
# Convert data to long form for the analyses, adding Time:
# important to order the data with respect to the subjects:
BtBL <- gather(BtB, key = Visit, value = BDI, BDI2m, BDI4m, BDI6m, BDI8m) %>%
  mutate(Time = as.integer(substr(Visit, 4, 4))) %>%
  arrange(Subject)
```

\pagebreak


## Table 10.3


```{r, echo=TRUE}
# independence correlation structure
BtB_gee1 <- geepack::geeglm(BDI ~ BDIpre + Time + Treatment + Drug + Length,
                     id = Subject, 
                     data = BtBL, 
                     family = gaussian, 
                     corstr = "independence", std.err="san.se")
```

```{r}
summary(BtB_gee1)
```

\pagebreak

## Table 10.4

```{r}
# exchangeable correlation structure
BtB_gee2 <- geepack::geeglm(BDI ~ BDIpre + Time + Treatment + Drug + Length,
                     id = Subject, 
                     data = BtBL, 
                     family = gaussian, 
                     corstr = "exchangeable", std.err="san.se")
```

```{r}
summary(BtB_gee2)
```

## Auto-regressive lag-one correlation structure

Not shown in the textbook

```{r}
BtB_gee3 <- geepack::geeglm(BDI ~ BDIpre + Time + Treatment + Drug + Length,
                     id = Subject, 
                     data = BtBL, 
                     family = gaussian, 
                     corstr = "ar1", std.err="san.se")
```

```{r}
summary(BtB_gee3)
```

## Unstructured correlation structure

Not shown in the textbook

```{r}
BtB_gee4 <- geepack::geeglm(BDI ~ BDIpre + Time + Treatment + Drug + Length,
                     id = Subject, 
                     data = BtBL, 
                     family = gaussian, 
                     corstr = "unstructured", std.err="san.se")
```

```{r}
summary(BtB_gee4)
```


\pagebreak

## References

Højsgaard, S. Halekoh, U., & Yan, J. (2006).
[The R Package geepack for Generalized Estimating Equations](https://www.jstatsoft.org/article/view/v015i02)
*Journal of Statistical Software*,
15(2), 1-11.

Sheu, C.-F. (2000). [Regression analysis of correlated binary outcomes.](https://link.springer.com/content/pdf/10.3758/BF03207794.pdf)
*Behavior Research Methods, Instruments, & Computers*, 
32(2), 269-273.

\pagebreak

## Session information

```{r session_info, include=FALSE}
# set include=TRUE to view the results
sessionInfo()
```

