---
title:  |  
  | *Multivariate Analysis for the Behavioral Sciences,*
  | **Examples of Chapter 13:**
  | **Principal Components Analysis**
author: |
  | Kimmo Vehkalahti, Brian S. Everitt; edited by C.-F. Sheu
  |
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  rmarkdown::github_document:
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  fig.align = 'center',
  comment = "",
  tidy = FALSE,     # display code as typed
  size = "small")   # slightly smaller font for code)
```


```{r message=FALSE, warning=FALSE}
# check to see if the pacman package is there
# if not install it and then use it manage packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, calibrate)
```


\pagebreak

## Example: Drug usage

Data on drug usage rates for 1,634 students in the seventh
to ninth grades in 11 schools in the greater metropolitan area of Los Angeles.
Each participant completed a questionnaire about the number of times 
(on a five-point scale)
a particular substance (out of 11) had ever been used.


## Table 13.8: Correlation Matrix for Drug Usage Data

```{r}
source("https://www.york.ac.uk/depts/maths/data/everitt/chap4druguse.dat")
druguse.cor
```

\pagebreak

## Table 13.9

```{r}
options(digits = 3)
drugs_pc <- princomp(covmat = druguse.cor)
print(drugs_pc)
summary(drugs_pc, loadings=TRUE)
```

\pagebreak

## Figure 13.7

```{r fig13.7}
# plot(1:13, drugs_pc$sdev^2, type = "l", 
#      xlab = "Component Number", ylab = "Variance")
screeplot(drugs_pc, type = "lines", 
          main = "Screeplot of Drug Use Data")
abline(h = 1, col = "indianred", lty = 3)
legend("topright", legend = "Eigenvalue = 1", bty = "n", 
       col = "indianred", lty = 3, cex = 0.7)
```

\pagebreak

## Session information

```{r session_info, include=FALSE}
# set include=TRUE to view the results
sessionInfo()
```
